---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FinanceGraphs

<!-- badges: start -->
<!-- badges: end -->

A flexible wrapper around [dygraphs](https://dygraphs.com/) and [ggplot2](https://ggplot2.tidyverse.org/) to graph and annototate financial time series data.  
The package provides several ways to add additional information to a simple series vs time data, including horizontal annotations (events highligted by 
lines and colored bands) and vertical annotations (key levels or regions).  Colors and set of labels can be customized and persist across invocations 
of the package, but sensible defaults are used. To minimize verbiage to get what you want, the package emphasizes options in the main function call 
rather than multiple functions or pipes to build up a graph.

You can install the development version of FinanceGraphs from 

```{r, eval=F}
pak::pak("derekholmes0/FinanceGraphs")
# install.packages("FinanceGraphs")  # If from CRAN
```

# Dygraphs for time series

## Input data and Constants

The package is designed to be flexible with input data.  The data can either be in **long** format (e.g. `date`,`series`,`value`) or **wide** format (`date`,`col1`,`col2`,...).  
The `date` column can be called anything or be anywhere in the input data.frame, but there must be at least one coercible column with dates and one
numeric column.  Each column after the date column is treated as a separate series to be graphed, but if the column name (series name) ends in any of
'`.lo`, `.hi`, `.f` (for forecast), '`.flo`, `.fhi` then those columns are treated specially as lower/upper bands or forecast series associated with 
the prefix of the name.

Consistent colors used for all series and annotations are kept in a local (persistent) settings `data.frame`.  Default color schemes can
be changed (again persistently) using the `fg_update_aes()` function.

## Simple Examples

```{r, echo=FALSE, message=FALSE, warning=FALSE}
options(datatable.print.keys=FALSE,datatable.print.class=FALSE)
require(data.table)
require(FinanceGraphs)
```

```{r}
#| label: simple_example1
#| fig.alt: "Simple Graph of a few Equity Prices"
fgts_dygraph(eqtypx, title="Stock Prices", ylab="Adjusted Close")
```
* All of the data is displayed on the graph by default, but nost of the time, we want to focus on recent periods and have the option of backing up to all history.
The `dtstartfrac` parameter shows just the last (1-`dtstartfrac`) percent of the data.  Alternatively, a *generic date window* of the form "start::end" can be 
used in the `dtwindow` parameter. Each end of the window can be a full date (e.g. "2022-01-01") or a relative date string (e.g. "-6m" for 6 months ago.

* One of the most appealing features of [dygraphs](https://dygraphs.com/) is the ability to smooth series interactively.  The `roller` parameter adds a rolling average smootherof the specified width (in data points).  A default smoothing parameter is chosen depending on the length of the underlying data, but can be overridden using the
`roller` parameter.

* Individual series can be altered in several ways:

  - Highlighted with different stroke patterns and width using the `hilightcols` argument with a string (possibly list) of series names, `hilightwidth` for a new width, and `hilightstyle` for a new stroke pattern.
  - Shown a step plot using the `stepcols` argument with a string (possibly list) of series names.
  - Hidden from using the `hidecols` argument with a string (possibly list) of series names.
  - Rebased to a given constant at a particular date.

```{r}
#| label: simple_example2
#| fig.alt: "Graph of Equity Prices showing highlights and date selection methods"
fgts_dygraph(eqtypx, title="W/ Focused range, highlights, rebasing",
             dtstartfrac=0.6,hilightcols="QQQ",hilightwidth=4,rebase="2024-01-01,100",roller=3)
```

* Series can be grouped together into bands by adding new columns in the data with names ending in '`.lo`' and '`.hi`' for lower and upper bounds. 
Those additional series can represent many things, such as statistical extremes, rolling correlations to other variables (see vignette),
or (as a special case) forecast confidence intervals.

* Horizontal annotations can also be added using the `annotations` parameter.  The most common example is a horizontal line at the
last observations of each series.  

```{r}
#| label: simple_example3
#| fig.alt: "Hi/Lo series, used to show currency values against local peers"
toplot <- reerdta[REGION=="LATAM",.(cop=sum(value*(variable=="COL")),
               peers=mean(value),peers.lo=min(value),peers.hi=max(value)),by=.(date)]
fgts_dygraph(toplot,title="COP REER vs Latam peers",ylab="Price",
             roller=1,hilightcols="cop",hilightwidth=4,annotations="last,linevalue")
```

## Events

Annotations to a particular date or date range can be added to the graph using the `events` and `event_ds` parameters.  
The `events` parameter is a string with one or more (separated by semicolons) event specifications.  The `event_ds` parameter is an optional data.frame with user defined events. (See Vignette for examples)
Any events specified with either parameter are additive. 

### Events in the function call

Events always have a start date and a text label.  THey are shown as vertical lines on the graph unless they also have an End Date, in which case the entire region is shaded.   Event strings starting with `doi` (date of interest) are predefined events included with the package.  Those are customizable using the `fg_update_dates_of_interest()` and can be listed using `fg_list_dates_of_interest()`.
Event strings can be added together with semicolons, as in the following example:

```{r}
#| label: Events1
#| fig.alt: "Graph of Equity Prices showing Event styles"
smalldta <- eqtypx[date>=as.Date("2023-01-01"),.(date,IBM,QQQ)]
fgts_dygraph(smalldta,title="With Events",ylab="Price",events="doi,regm;doi,fedmoves;date,xmas,2025-12-25")
```
Several types of events are predefined in `fgts_dygraph()` including equity option expirations, IMM CDS roll dates, seasonal events (e.g. "same day in quarter as last observation") and series extremes.  See the vignette for more examples.  Events can also be passed in as a `data.frame` using the `event_ds` parameter, as shown next.

### Event helpers

Custom events can also be passed in `data.frame` format using the `event_ds` parameter.  The details are in `fgts_dygraph()`, but the
basic columns are a start date date, a text label, and if applicable an end date. Only events within the dates of the original input
data are shown.

The package includes a few "event helpers" to make it easy  to generate the right formats for given more complicated types of events and exogenous data.  See the vignette for examples, but here is an idea of two that can be done.

```{r}
#| label: Events2
#| fig.alt: "Graph of equity prices showing sentiment ranges"
events_consumer_sent <- fg_cut_to_events(consumer_sent,center="zscore")
head(events_consumer_sent,2)
fgts_dygraph(smalldta,title="Equity Prices w Sentiment",event_ds=events_consumer_sent)
```
Current event helpers are:

|Function|Description|
|:-------|:-----------|
|`fg_findTurningPoints()`|  Statistically identify turning points in a series|
|`fg_ratingsEvents()`|  Add colored ranges based on analyst credit ratings|
|`fg_cut_to_events()`| "Cut" a univariate series into colored bands, with two different colors for positive and negative values|
|`fg_signal_to_events()`| Map a long/short signal to events |
|`fg_tq_divs()`|Add dividend events from TidyQuant dividend data|
|`fg_av_earnings()`|Add earnings events from AlphaVantage earnings data|

## Forecasts

Forecasts beyond the last day of the dataset can also be added in a consistent way.  For example, forecasts (and confidence intervals)
for the IBM stock price in new `data.frame` can be added as follows.  Each forecast is shown as the same color as the original
series, but dashed to show the transition.

```{r}
#| label: Forecasts1
#| fig.alt: "Adding forecasts to original data"
head(example_fcst_set,2)
fgts_dygraph(smalldta,title="With Forecasts", dtstartfrac=0.7,forecast_ds=example_fcst_set)
```

Like events, forecasts can be generated from many packages with different output formats.  There is also a "forecast helper" to
get their outputs into the appropriate `foreast_ds` forms

|Function|Description|
|:-------|:-----------|
|`fg_sweep()`|Converts [tidy forecast objects](https://business-science.github.io/sweep/) to `forecast_ds` form.|
|`fg_prophet()`|Converts [Prophet](https://business-science.github.io/sweep/) to `forecast_ds` form.|

## Changing aesthetics and adding dates of interest

Default colors for series and annotations can be changed using `fg_update_aes()` or `fg_update_line_colors()`.
Any changes made to aesthetics colors will persist across loads of the package (unless `persist=FALSE` is specified).
To see what aesthetics are used for any particular function, use `fg_print_aes_list()`
As an example, to make a graduated set of colors for the first 2 series.

```{r}
fg_get_aes("lines",n_max=2)
fg_update_line_colors( rev(RColorBrewer::brewer.pal(8,"GnBu"))[1:2] )
fg_get_aes("lines",n_max=3)
```

New dates of interest used for the `events` parameter can also be added.  To add (for example) a new FOMC cut of 50bps on 6/16/2026
(after the development of this package), use `fg_update_dates_of_interest()`.  Resetting the lists (and colors) can also be done.

```{r}
newdoi <-data.frame(category="fedmoves",eventid="F:-50",DT_ENTRY=as.Date("6/16/2026",format="%m/%d/%Y"))
fg_update_dates_of_interest(newdoi)
tail(fg_get_dates_of_interest("fedmoves"),2)  |> data.frame()
fg_reset_to_default_state()
```

## Integration into Markdown and Shiny

Dygraphs have the very nice feature of allowing synchronized zoom in Markdown or Shiny applications.  Each graph with a common `group` identifier
is synchronized.  There are times, however, when you want to turn these on or off.  The function `fg_sync_group()` either returns the current group name
if called with no parameters, sets the group name with a string, or turns off synchronization with a call `fg_sync_group(NULL)`.

# Static Plots for Time Series

## Scatter Plots with time dimension enhancements

Key to understanding how time series co-move is a simple scatter plot.  The function `fg_scatplot()` tries to be a concise wrapper around the very comprehensive
[ggplot2](https://ggplot2.tidyverse.org/) graphics framework.  `ggplot2()]`` is a great ecosystem, but requires quite a bit of verbiage to get from idea to presentable graph quickly.  The approach used here is to specify broad categories of aesthetics with a formula, while the details are kept behind the hood using the aesthetic sets managed by `fg_get_aes()` as above.  Fuller explanations and more examples are in the accompanying vignette.

This "one-line" approach can be used with both date-based and non-date based datasets.  For example, suppose we wanted to plot two asset prices against
each other so that we can easily understand both how they co-move, where they have been recently, and where they are now.  We will start by making a
fake dataset of two sets of two assets each. 

```{r}
set.seed(1)
ndates <- 400;
samp_rw <- function() { 100*(1+cumsum(rnorm(ndates,sd=0.2/sqrt(260)))) }
dts <- seq(as.Date("2021-01-01"),as.Date("2021-01-01")+ndates-1)
dttest <- rbind( data.table(date=dts,ccat="A",px_x=samp_rw(),px_y=samp_rw()),
                 data.table(date=dts,ccat="B",px_x=samp_rw(),px_y=samp_rw()))
```

A scatter plot of where those two assets are and where they have been is as easy as specifying (in one string!) columns to plot, a column for
color, and a term to split the dates into 2 month and 6 month intervals, and finally a term to show where the latest observations are.
As with `fgts_dygraph()`, date partitions can be custom regimes managed by `fg_update_dates_of_interest()`.

```{r}
#| label: Scatter1
#| fig.alt: "Scatterplot with two categories and recent highlighting."
fg_scatplot(dttest,"px_y ~ px_x + color:ccat + doi:recent + point:label","lmone",datecuts=c(60,182),title="Splitting dates")
```
One important feature of the static plots in this package to note is that they include options for switching the colors of large numbers of groups
or points from discrete values to continuous values from the [RColorBrewer](https://colorbrewer2.org) package.

`fg_scatplot()` does not require dates in the input data, and can be useful for communicating comparative analyses as well.  The `mtcars` dataset
shows how we can add a lot of information to a basic text scatterplot.  

```{r}
#| label: Scatter2
#| fig.alt: "Scatterplot with text and linear models"
dt_mtcars=data.table(datasets::mtcars)[,let(id=lapply(rownames(datasets::mtcars),\(x) last(strsplit(x," ")[[1]])))]
fg_scatplot(dt_mtcars,"disp ~ hp + color:carb + label:id","lmonenoeqn",n_color_switch=0,title="Text with color switch")
```

Many more examples that encapsulate a large part of the [ggplot2](https://ggplot2.tidyverse.org/) corpus are in the accompanying vignette.

## Time-categorized box plots

One way to visualize multiple time series comparatively with a boxplot or a violin plot.  The function `fg_tsboxplot()` provides a flexible way
to split several time series into time categories.  Usually, those categories are time periods with increasing horizons, but this function
allows for arbitrary periods managed by `fg_update_dates_of_interest()`.  Data can be normalized across historical categories and time series
categories.  As an example, suppose we would like to see normalized equity prices over the past two years, split into categories of the last
week, month (less the last week) and the last quarter.  

```{r}
#| label: Boxplot1
#| fig.alt: "Boxplot of relative equity prices"
fg_tsboxplot(narrowbydtstr(eqtypx,"-2y::"),breaks=c(7,90),normalize="byvar",title="Normalized Equity prices by Date category")
```

Like `fgts_dygraph()`, the function can take data in both wide and long formats.  Longer formats have the advantage of adding more categorical levels 
to our graphs.  For example, we can plotting recent ranges of the monthly FX Real Exchange Rate dataset in the following way.
Here, we split the data into buckets of last 20pct of the months, the first half the months and the last. We will also show the last 
observation, hide the boxplot whiskers, and reorder the currencies by their relative weakness.

```{r}
#| label: Boxplot2
#| fig.alt: "Boxplot of Real Effective Exchange Rates"
fg_tsboxplot(reerdta,breaks=c(0,0.2,0.5,1),doi="last",orderby="value",boxtype="nowhisker",facetform=". ~ REGION",title="Real Eff. Exch Rates")
```

## Event Studies

Trying to understand how events or environments may impact prices can be summarized well by plotting their movements relative to a date
against time relative to the event.  The function `fg_eventStudy()` integrates a set of dates with a set of time series to plot
their relative behavior over the business days before and after each event.  It is designed to show reasonably large sets of 
assets or event dates by switching discrete color scales to gradient scales after a user specified number.

There are several ways to show the data, including path-by-path, statistics by event (i.e. across assets) or by asset (across events), 
boxplots, or scatter plots of moves at the edge of the intervals.

For example, to see the behavior of various parts of the US yield curve around FOMC cuts, just use the following

```{r}
#| label: EventStudies1
#| fig.alt: "Relative rate moves around recent Fed Cuts"
dtset <- fg_get_dates_of_interest("fedmoves")[grepl("F:-",eventid),.(DT_ENTRY,text=eventid2)]
fg_eventStudy(yc_CMSUST,dtset,output="pathbyvar",
              title="Constant Maturity UST tenors around Fed Cuts")
```

We can see that curves reach their peak steepening around a week after the event.  To see this a little more succinctly, we can use

```{r}
#| label: EventStudies2
#| fig.alt: "Scatter plot of moves before and after Fed Rate Cuts"
fg_eventStudy(yc_CMSUST,dtset,output="scatter",nbd_back=5,nbd_fwd=6,
                title="Constant Maturity UST tenors around Fed Cuts at 5 days")
```

















