---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FinanceGraphs

<!-- badges: start -->
<!-- badges: end -->

A flexible wrapper around [dygraphs](https://dygraphs.com/) and [ggplot2](https://ggplot2.tidyverse.org/) to graph and annototate financial time series data.  
The package provides several ways to add additional information to a simple series vs time data, including horizontal annotations (events highligted by 
lines and colored bands) and vertical annotations (key levels or regions).  Colors and set of labels can be customized and persist across invocations 
of the package, but sensible defaults are used. To minimize verbiage to get what you want, the package emphasizes options in the main function call 
rather than multiple functions or pipes to build up a graph.

You can install the development version of FinanceGraphs from 

```{r, eval=F}
pak::pak("derekholmes0/FinanceGraphs")
# install.packages("FinanceGraphs")  # If from CRAN
```


# Dygraphs for time series

## Input data and Constants

The package is designed to be flexible with input data.  The data can either be in **long** format (e.g. `date`,`series`,`value`) or **wide** format (date,col1,col2,...).  
The `date` column can be called anything or be anywhere in the input data.frame, but there must be at least one coercible column with dates and one
numeric column.  Each column after the date column is treated as a separate series to be graphed, but if the column name (series name) ends in any of
'`.lo`, `.hi`, `.f` (for forecast), '`.flo`, `.fhi` then those columns are treated specially as lower/upper bands or forecast series associated with 
the prefix of the name.

Consistent colors used for all series and annotations are kept in a local (persistent) settings `data.frame`.  Default color schemes can
be changed (again persisantly) using the [fg_update_colors()] function.

## Simple Examples

```{r}
#| label: simple_example1
#| fig.alt: "Simple Graph of a few Equity Prices"
require(FinanceGraphs)
fgts_dygraph(eqtypx, title="Stock Prices", ylab="Adjusted Close")
```
* All of the data is displayed on the graph by default, but nost of the time, we want to focus on recent periods and have the option of backing up to all history.
The `dtstartfrac` parameter shows just the last (1-`dtstartfrac`) percent of the data.  Alternatively, a *generic date window* of the form "start::end" can be 
used in the `dtwindow` parameter. Each end of the window can be a full date (e.g. "2022-01-01") or a relative date string (e.g. "-6m" for 6 months ago.

* One of the most appealing features of [dygraphs] is the ability to smooth series interactively.  The `roller` parameter adds a rolling average smoother 
of the specified width (in data points).  A default smoothing parameter is chosen depending on the length of the underlying data, but can be overridden using the
`roller` parameter.

* Individual series can be altered in several ways:

  - Highlighted with different stroke patterns and width using the `hilightcols` argument with a string (possibly list) of series names, `hilightwidth` for a new width, and `hilightstyle` for a new stroke pattern.
  - Shown a step plot using the `stepcols` argument with a string (possibly list) of series names.
  - Hidden from using the `hidecols` argument with a string (possibly list) of series names.
  - Rebased to a given constant at a particular date.

```{r}
#| label: simple_example2
#| fig.alt: "Graph of Equity Prices showing highlights and date selection methods"
fgts_dygraph(eqtypx, title="W/ Focused range, highlights, rebasing",
             dtstartfrac=0.6,hilightcols="QQQ",hilightwidth=4,rebase="2024-01-01,100",roller=3)
```

* Series can be grouped together into bands by adding new columns in the data with names ending in '`.lo`' and '`.hi`' for lower and upper bounds. 
Those additional series can represent many things, such as statistical extremes, rolling correlations to other variables (see vignette),
or (as a special case) forecast confidence intervals.

* Horizontal annotations can also be added using the `annotations` parameter.  The most common example is a horizontal line at the
last observations of each series.  

```{r}
#| label: simple_example3
#| fig.alt: "Hi/Lo series, used to show currency values against local peers"
toplot <- reerdta[REGION=="LATAM",.(cop=sum(value*(variable=="COL")),
               peers=mean(value),peers.lo=min(value),peers.hi=max(value)),by=.(date)]
fgts_dygraph(toplot,title="COP REER vs Latam peers",ylab="Price",
             roller=1,hilightcols="cop",hilightwidth=4,annotations="last,linevalue")
```

## Events

Annotations to a particular date or date range can be added to the graph using the `events` and `event_ds` parameters.  
The `events` parameter is a string with one or more (separated by semicolons) event specifications.  The `event_ds` parameter is an optional data.frame with user defined events. (See Vignette for examples)
Any events specified with either parameter are additive. 

### Events in the function call

Events always have a start date and a text label.  THey are shown as vertical lines on the graph unless they also have an End Date, in which case the entire region is shaded.
Event strings starting with `doi` (date of interest) are predefined events included with the package.  Those are customizable using the [fg_update_dates_of_interest()] and can be listed using [fg_list_dates_of_interest()].
Event strings can be added together with semicolons, as in the following example:

```{r}
#| label: Events1
#| fig.alt: "Graph of Equity Prices showing Event styles"
smalldta <- eqtypx[date>=as.Date("2023-01-01"),.(date,TLT,EEM)]
fgts_dygraph(smalldta,title="With Events",ylab="Price",events="doi,regm;doi,fedmoves;date,xmas,2025-12-25")
```
Several types of events are predefined in [fgts_dygraph()] including equity option expirations, IMM CDS roll dates, seasonal events (e.g. "same day in quarter as last observation") and series extremes.
See the vignette for more examples.  Events can also be passed in as a `data.frame` using the `event_ds` parameter, as shown next.

### Event helpers

Custom events can also be passed in `data.frame` format using the `event_ds` parameter.  The details are in [fgts_dygraph()], but the
basic columns are a start date date, a text label, and if applicable an end date. Only events within the dates of the original input
data are shown.

The package includes a few "event helpers" to make it easy  to generate the right formats for given more complicated types of events and exogenous data.  See the vignette for examples, but here is an idea of two that can be done.

```{r}
#| label: Events2
#| fig.alt: "Graph of equity prices showing sentiment ranges"
events_consumer_sent <- fg_cut_to_events(consumer_sent,center="zscore")
head(events_consumer_sent,2)
fgts_dygraph(smalldta,title="Equity Prices w Sentiment",event_ds=events_consumer_sent)
```
Current event helpers are:

|Function|Description|
|:-------|:-----------|
|`fg_findTurningPoints()`|  Statistically identify turning points in a series|
|`fg_ratingsEvents()`|  Add colored ranges based on analyst credit ratings|
|`fg_cut_to_events()`| "Cut" a univariate series into colored bands, with two different colors for positive and negative values|
|`fg_signal_to_events()`| Map a long/short signal to events |
|`fg_tq_divs()`|Add dividend events from TidyQuant dividend data|
|`fg_av_earnings()`|Add earnings events from AlphaVantage earnings data|

## Forecasts

Forecasts beyond the last day of the dataset can also be added in a consistent way.  For example, forecasts (and confidence intervals)
for the IBM stock price in new `data.frame` can be added as follows.  Each forecast is shown as the same color as the original
series, but dashed to show the transition.

```{r}
#| label: Forecasts1
#| fig.alt: "Adding forecasts to original data"
smalldta <- eqtypx[date>=as.Date("2023-01-01"),.(date,IBM,QQQ)]
head(example_fcst_set,2)
fgts_dygraph(smalldta,title="With Forecasts", dtstartfrac=0.7,forecast_ds=example_fcst_set)
```

Like events, forecasts can be generated from many packages with different output formats.  There are also "forecast helpers" to
get their outputs into the appropriate `foreast_ds` forms

|Function|Description|
|:-------|:-----------|
|`fg_sweep()`|Converts [tidy forecast objects](https://business-science.github.io/sweep/) to `forecast_ds` form.|
|`fg_forecast()`|Converts [forecasts](https://pkg.robjhyndman.com/forecast/) to `forecast_ds` form.|

## Changing colors and adding dates of interest

Default colors for series and annotations can be changed using [fg_update_colors()].  Any changes made to colors will persist across
loads of the package.  As an example, to make a graduated set of colors for the first 5 series.

```{r}
fg_get_colors("lines",n_max=3) -> oldcolors
fg_update_colors( oldcolors[,let(color=  rev(RColorBrewer::brewer.pal(8,"GnBu"))[1:3])][] )
fg_get_colors("lines",n_max=3)
```

New dates of interest used for the `events` parameter can also be added.  To add (for example) a new FOMC cut of 50bps on 6/16/2026
(after the development of this package), use [fg_update_dates_of_interest()].  Resetting the lists (and colors) can also be done.

```{r}
newdoi <-data.table(category="fedmoves",eventid="F:-50",DT_ENTRY=as.Date("6/16/2026",format="%m/%d/%Y"))
fg_update_dates_of_interest(newdoi)
tail(fg_get_dates_of_interest("fedmoves"),2)  |> data.frame()
fg_reset_to_default_state()
```

