---
title: "FinanceGraphs-Customization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FinanceGraphs-Customization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(FinanceGraphs))
```

```{r,echo=FALSE}
options(datatable.print.keys=FALSE, datatable.print.class=FALSE)
knitr::opts_chunk$set(fig.width = 7)
```

# Overview of Aesthetics Organization

Managing a consistent look across graphs is not easy, as there are so many parameters that are possible
to change.   [ggplot2](https://ggplot2.tidyverse.org/) does a great job allowing every detail to be
customized, especially with the use of themes.  However, adding all those customizations are
burdensome, and ad-hoc changes to them can involve a great deal of programming.  Ideally, we would like
to keep our favorite looks in one place with an easy ability to change just a few pieces in an 
ad-hoc manner.

The functions in the package attempt to ease that burden with a *middle layer* of named aesthetic groups.
Aesthetics can be managed just like any other data, so internal to the package is a `data.table` with 
the following character columns, plus a few others:

|Column|Description|
|:-------|:------------------------|
|Category|An "aesthetic set" key grouping together various subparts of a graph.|
|(Ordering) Variable|Any string that can sorted to keep the aesthetics in a desired order|
|Type|The type of aesthetic (e.g. color, size, symbol, etc)
|Value|The actual aesthetic to be used.  If a number, then each function casts appropriately|
|used|Where the function is used|

Each FinanceGraphs function has internal aesthetic names (and defaults) that a user can modify, either
persistently or temporarily.  For example, the basic set of colors for categories is `"lines"`, which
are the line colors for `fgts_dygraph()` or the category colors for `fg_scatplot`.

```{r}
fg_get_aes("lines", n_max=6)
```

Changing the look of graphs is as easy as changing the values in that internal dataset.  In some cases
(e.g. in `fg_scatplot()`) a user can make an entire new aesthetic set with a new (unique) name and directly
specify it in the function call.  The package keeps those changes persistently (by default), so each
users' preferences should only need to be specified once.

# Getting current aesthetics

There are two ways to get the aesthetic sets for every graph in the package:

* Call the function `fg_print_aes_list()` with the name of a function as an argument
* Call the function `fg_verbose()` to turn on logging of what aesthetic sets are called for any given
graph generated.

For example, here are the a few aesthetic sets for `fg_eventStudy()`. This function just produces a summary,
so for example there are actually 14 different line colors in `"lines"`.

```{r}
print(fg_print_aes_list("fg_eventStudy"))
```

Each set can have multiple rows, as in 

```{r}
fg_get_aes("espath_ls", n_max=6)
```
These correspond directly to [ggplot2](https://ggplot2.tidyverse.org)  aesthetics, such as the linetypes in [linetypes](https://ggplot2.tidyverse.org/reference/aes_linetype_size_shape.html#linetype).  Some other notes:

* Each value is a character column, but is coerced as necessary in each function.  
* In the cases of sizes, the values are multiples of the function parameters `psize` and `tsize`. So, a text size
value of `"6"` gets converted to `tsize * 6` in each function.
* The `"variable"` column tells the functions to use `"L01"` first, `"L02"` second, etc.  As long as the key is sortable,
consistent and correct values should be used. 
* In the case of category overload, most functions have "color switch" parameters beyond which colors are "brewed" scales
which are infinitely (but not wisely) expandable.  In this case, the value is `"seq,<colorset>"` which generates colors
from a [Colorbrewer](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) color set.
* Colors can be color names or Hex color codes.

Colors in particular may be examined using the `fg_display_colors()` function as in

```{r}
fg_display_colors("lines")
```

# Adding or customizing aesthetics

Any of the default aethetic sets can be customized across calls to the functions and invocations of the package using `fg_update_aes()`   New aesthetics sets can also be added for those functions (e.g. `fg_scatplot()`) where different
aesthetic sets can be specified at runtime.

Modifying or adding aesthetics sets is done by creating (or copying and editing) a `data.frame` obtained from `fg_get_aes()`
As a simple example, suppose we have three related classes of assets, one of which we wish to highlight and the others are
related, but less important. Here is how the default colors would look:

```{r}
onedt <- function(offset,category) { data.table(x=seq(1,10),y=seq(1,10)-offset,id=rep(category,10))}
exampledta <- rbind(onedt(0,"A"),onedt(1,"B"),onedt(2,"C"))
fg_scatplot(exampledta,"y ~ x + color:id",title="default",psize=6)
```

## Customizing default aesthetics

The procedure for changing those colors is as follows:

1. Get an example `data.frame` using `fg_get_aes()`
2. Modify the values you want to change in the `value` column
3. Update the persistent package data using  `fg_update_aes()`

```{r}
head(oldcolors <- fg_get_aes("lines"),3)
oldcolors[c(1,2,3),"value"] <- alpha("blue", c(1,0.6,0.3))
# Note that we still keep "category" as "lines".  To add a new set, use a different name.
fg_update_aes( oldcolors )
fg_display_colors("lines")
```

```{r}
fg_scatplot(exampledta,"y ~ x + color:id",title="default",psize=6)
```

## Adding new aesthetics

To create our own aesthetics, we use the same procedure, but adding our own `"category"`:

```{r}
oldcolors[c(1,2,3),"value"] <- alpha("red", c(1,0.6,0.3))
oldcolors[c(1,2,3),"category"] <- rep("MyNewColors",3)
fg_update_aes( oldcolors )
fg_scatplot(exampledta,"y ~ x + color:id,MyNewColors",title="with new colors",psize=6)
```

# Themes

Themes are the [ggplot2](https://ggplot2.tidyverse.org) way of proscribing every single aesthetic
detail in a graph.  This package uses a default theme derived from `theme_bw()`,
but it is quite easy to create or modify, and more importantly **save**, a custom theme for future use.

To do so, just call `fg_replace_theme()` as in the following example:

```{r}
fg_replace_theme(theme_dark())
fg_scatplot(exampledta,"y ~ x + color:id,MyNewColors",title="New Theme, New Colors",psize=6)
```

# Persistence

## Saved Aesthetics, themes and dates of interest persist

This package manages aesthetic changes for you by caching the current aesthetic sets, themes,
and dates of interest in local files, which are then loaded on package invocation. If you **don't**
want save changes, then call `fg_update_aes()` and `fg_replace_theme()` with `persist=FALSE` parameters.

To reset all parameters back to the package defaults, run `fg_reset_to_default_state()`




