---
title: "Time-Series-scatterplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-Series-scatterplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(stringr))
suppressPackageStartupMessages(require(FinanceGraphs))
```

```{r,echo=FALSE}
options(datatable.print.keys=FALSE, datatable.print.class=FALSE)
knitr::opts_chunk$set(fig.width = 7)
```

# Scatter Plots using `fg_scatplot()`

Scatter plots are one of the most useful ways of identifying relationships 
in financial data.  Typically the plots are of two variables where the data
points are ordered by time, which cannot easily be communicated statically.
The goal of this function is to create graphs that bring time back into
the graph, as well as give a new system for quickly getting the results desired.

R graphing packages are incredibly flexible and robust, but to really use them
well requires (1) a learning curve for many different functions and parameters, and 
(2) necessarily a lot of code to get anything other than the simplest of graphs.

Distilling all those programmatic features can be done reasonably easily, but at
the expense of almost as many paramters into a single wrapper function.  For years,
that was my approach, but here is an even more succinct approach:  **Customize graphs
with simple formulas**.

## Describing components of scatter plot formulas

In addtion to the basic `y ~ x` needed to identify the two variables to plot, additional
customization can be added by associating a column in the data set with each customization.
Not all customizations require a column, and sometimes we might want to add additional
information to the customization (aesthetic).  The general format is:

`y ~ x + <feature_1>:<column>,<aesthetic group> + <feature_2>:... + ... `

The best way to see this is with a few examples of features using the well-known `mtcars` data set.  
Suppose we want to plot displacement `disp` vs horsepower `hp`, but we want also to
be able to identify individual cars and color the labels by number of cylinders `cyl`.
Suppose we already have added the labels as a column `id` as below. 

```{r, echo=FALSE}
mtcars=data.table(datasets::mtcars,keep.rownames=TRUE)
mtcars=mtcars[,id:= stringr::str_split_i(rn," ",-1)]
```

```{r}
head(mtcars,2)
fg_scatplot(mtcars,"disp ~ hp + color:cyl + label:id","scatter",title="mtcars")
```

The next sections describe the features that can be added as terms to the
input formula, by general category.

### Aesthetic terms in plot formula

Each term is of the general form `aesthetic:<column>,<aestheticset>` where the
(not always required) column is used to determine the levels of each aesthetic.
`<aesthetic set>` is described at the end of the vignette.  Individual points on a 
graph can always either just be points or symbols, or have a text annotations 
per data point.  Points can be distinguished by

|Term|Parameters|Description|
|:-----|:----------|:------------------|
|`color`|`colname,<aesset>`|Color of each point or label from **levels** of `colname`|
|`size`|`colname,<aesset>`|Size of each point from **levels** of `colname`|
|`symbol`|`colname,<aesset>`|Symbol of each point from **levels** of `colname`|

To create text, you can use

|Term|Parameters|Description|
|:-----|:----------|:------------------|
|`text`|`colname,<aesset>`|Text in character column `colname` at each `x,y` point.|
|`label`|`colname,<aesset>`|Bordered label in character column `colname`|
|`labelhilight`|`colname,<aesset>`|Filled in (and bordered) label|
|`tooltips`|`colname`|Plot points, but with mouseover labels (1)|

(1): Refer to [ggiraph](https://davidgohel.github.io/ggiraph/) for details.  Note that output 
from `fg_scatplot` must be displayed using the `girafe()` function.

An admittedly too complex example of combining these together is

```{r}
fg_scatplot(mtcars,"disp ~ hp + color:cyl + symbol:gear + size:am","scatter",title="mtcars")
```

### Time Series related terms in the plot formula

Fortunately, there are many aesthetics with which we can use to understand evoluation through time.  
This package uses size, with larger points corresponding to more recent data. This choice allows for 
multiple relationships to be shown at once.  The user specifies how to partition the data (from the 
last point) using the `datecuts` parameter, and a `doi` (for Dates Of Interest) parameter.  
For the time based graphs, we first start with a simulated set of prices for two equities and an index:

```{r}
set.seed(1); ndates<-400
samp_rw <- function() { 100*(1+cumsum(rnorm(ndates,sd=0.2/sqrt(260)))) }
dts <- seq(as.Date("2021-01-01"),as.Date("2021-01-01")+ndates-1)
dttest_A <- data.table(date=dts,ccat="A vs X",px_idx=samp_rw(),px_eq=samp_rw())
dttest_B <- copy(dttest_A)[,let(ccat="B vs X",px_eq=samp_rw())]
dttest <- rbind(dttest_A, dttest_B)
```

The following
example compares two assets `A` and `B` against an index and puts an label at the last data points in the set.


```{r}
fg_scatplot(dttest,"px_eq ~ px_idx + color:ccat + doi:recent + point:label","scatter",
                datecuts=c(66,122),title="from recent")
```

All of the equation terms applicable to data sets with dates are:

|Term|Parameters|Description|
|:-----|:----------|:------------------|
|`doi`|`recent`|Partitions each time series by dates from the last point, using `datecuts`|
|`doi`|`<doiset>`|Partitions each time series into date ranges specified by `fg_update_dates_of_interest()`|
|`point`|`<value|label><all>`| Labels the last date by (x,y) coordinates or label|
|||`all` can be added to label last observations in each color category`|
|`point`|`<anno><all>`| Shows last values along `x` and `y` axes|

### Other terms in the plot formula

|Term|Parameters|Description|
|:-----|:----------|:------------------|
|`ellipse`||Add a equal bivariate frequency ellipse|
|`hull`|`<:quantile>`|Add the convex hull of points after taking out `quantile` points from center|
|`xline`|`<:x><,color>`|Adds a vertical line to graph at `x`|
|`yline`|`<:y><,color>`|Adds a horizontal line to graph at `y`|
|`grid`|`<dotted|dotted_x|dotted_y|none>`|Style of background grids|

## Graph types

Graph types are specified in the required `type` parameter, and control what additional 
`stats` to show along with a simple scatter (or density plot).  With a few exceptions,
the graph type is composed of two parts added together in a string, (1) the style that
points will be shown and (2) additional `stats` to statistically summarize the data.
Point styles are 

|Graph Type|Description|
|:-----|:----------|:------------------|
|`scatter`|Just plot `(x,y)` points or binned hexagons|
|`density`|Plot points as unfilled density plot.
|`path`|Plot points joined together sequentially|

For `density` or `scatter` point styles, summary regressions can be added with the following 
modifiers.  Note that just specifying the modifiers themselves implies a scatter plot
if possible.

|Graph Type|Optional Parameters|Description|
|:-----|:----------|:---------------------|
|`lm`||Add linear regression lines using `tformula` per category (2)|
|`loess`||Add Loess best fits|
||`<one>`|Add one linear or loess regression line using all data|
||`<noeqn>`|Suppress showing the resulting fits|
||`<nofill>`|Suppress confidence banks from shown regression lines|

(2): Levels used are the first among (`color`,`symbol`,`size`,`alpha`) specified in the `plotform`
formula.

As a more complex example of how these can be put together, suppose we want to find out if `EEM` and `IBM`
have different non-linear betas to `QQQ`.  First we use a poor-man's pivot, then

```{r}
dtrtn<- rbind(eqtyrtn[,.(date,r_eq=100*EEM,r_idx=100*QQQ,ticker="EEM")],
                    eqtyrtn[,.(date,r_eq=100*IBM,r_idx=100*QQQ,ticker="IBM")]) |>
                    narrowbydtstr("-1y::")
fg_scatplot(dtrtn,"r_eq ~ r_idx + color:ticker + xline:0 + yline:0","densitylm",
            tformula="y~0+x:(x>0)",title="piecewise linear")
```

The `path` type is useful to understand the evolution of two time series.  For example, the past year
of Stocks vs Consumer Sentiment can be seen with

```{r}
toplot = eqtypx[data.table(consumer_sent),on=.(date),roll=T] |> tail(n=12)
fg_scatplot(toplot,"QQQ ~ price + point:value","path",title="QQQ vs Sent",axislabels="ConsSent;Stonks")
```

## Bounding Boxes and other details

### Limiting the view port with bounding boxes

Many times there's always an outlier in Financial Time Series,  especially in Credit trading.
Outliers distort the graph, but removing them needs to be done with some care.  
`fg_scatplot()` has three ways to deal with outliers.  BY default, (1) all data is shown, but a 
"bounding box" can also be specified to narrow the view to the most relevant data.
Data outside that box can be either be (2) omitted, or (3) the preferred option of showing the data
at the edge of the box but with a clear notation that it lies somewhere outside the box.

The two parameters to control this are `boundboxtype` and `boundbox`.  The bounding box
can be specified as either actual values of each axis, or quantiles of the data along each axis.
The options for the bounding box type are 

|boundboxtype|Description|
|:-----|:----------------------|
|`value`|Omit any points outside the values of the bounding box|
|`valueidentify`|Squish the points into the box, noting if they are|
|`prob`|Omit any points outside the specified quantiles of the data|
|`probidentify`|Squish the points into the box, noting if they are|

Bounding boxes can either be lists of 2 or 4 numbers.  A two digit list truncates both axes equally,
while a 4 digit list truncates both lower and upper boxes of the data.  More explicitly, the
possibilities are:

|`boundboxtype`|`boundbox`|Description|
|:-----|:----------|:------------------|
|`value`|`c(y_min,y_max)`|x axis is unrestricted, y limited to [`y_min`,`y_max`]|
|`value`|`c(x_min,x_max,y_min,y_max)`|x axis limited to [`x_min`,`x_max`], y limited to [`y_min`,`y_max`]|
|`prob`|`c(q_x,q_y)`|x axis limited to `[q_x,1-q_x]` quantiles, y to `[q_y,1-q_y]`|
|`prob`|`c(q_lx,q_ux,q_ly,q_uy)`|x axis limited to `[q_lx,q_ux]` quantiles, y to  `[q_ly,q_uy]`|

The safest option in terms of seeing all the data is to have no bounding box,  but the next safest is to
use `probidentify`, as shown below.  

```{r}
fg_scatplot(mtcars,"disp ~ hp + text:id","scatter",title="boundbox Example",
             boundboxtype="probidentify",boundbox=c(0.1,0.1))
```

### Other annotations

Graphs sometimes require a lot of thought to understand what are the implications shown by the data.
One way to ease that communication is by adding notes (annotations) to the four corners of the graph
using the `annotatecorners` parameter.  You can also add an annotation to the x axis using
the semi-color separated parameter `xlabeldecoration` as shown below.

```{r}
fg_scatplot(dttest,"px_eq ~ px_idx + color:ccat ","scat",title="with decorations",
                    annotatecorners="NW;NE;SE;SW",
                    xdecoration="mktsad;mkthappy",ydecoration="eqsad;eqhappy")
```

### Faceting

`fg_scatplot()` renames columns internally and uses only the columns it really needs in the production
of the plot.  However, there are times when a user would like to keep columns in the original data with
the `ggplot` object.  This is particularly necessary if any further faceting is desired.  Faceting
columns can be added using the `keepcols` parameter as in

```{r}
require(ggplot2)
fg_scatplot(dttest,"px_eq ~ px_idx + point:value","scat",keepcols="ccat") + facet_wrap(ccat ~ .)
```


## Aesthetic customization

Managing a consistent look across graphs is not easy, as there are so many parameters that are possible
to change.   [ggplot2](https://ggplot2.tidyverse.org/) does a great job allowing every detail to be
customized, especially with the use of themes.  However, adding all those customizations are
burdensome, and ad-hoc changes to them can involve a great deal of programming.

The functions in the package attempt to ease that burden with a *middle layer* of named aesthetic groups.
Internally, there is a dataset that can be accessed with the function `fg_get_aes()` and 
managed with `fg_update_aes()`.  (See the accompanying vignette for more detail.)
For example, the **default** colors of points used in `fg_scatplot()` are the same as those used by the
lines in `fgts_dygraph()` and are taken fro the "lines" aesthetic set:

```{r}
fg_get_aes("lines",n_max=3)
```
The list of aesthetic sets used internally in each function can be seen by running `fg_print_aes_list()`
to return the names used internally and explanations.  To see what aesthetic sets are used for any given
plot, turn on verbosity using `fg_verbose()`. There are 23 categoeies used in `fg_scatplot()`, and the 
first five are:

```{r}
fg_print_aes_list("fg_scatplot") |> head(n=5)
```

Any of these can be customized across calls to the functions and invocations of the package using `fg_update_aes()`
You can also add new sets you might wish to use independently and then use them in invidual function calls
with the `aesset` added as in the options above.  For example, in the default aesthetic set there's
one called `"altlines_6"` which is shown below with an example of how to use it. Note that there are enough 
observations to kick the point display to the binned format.

```{r}
head(fg_get_aes("altlines_6"),2)
fg_scatplot(dttest,"px_eq ~ px_idx + color:ccat,altlines_6 + point:label","scatter",title="from recent")
```

User-customized themes are also always possible.  You can just add an alternte theme directly onto
the call such as `fg_scatplot(...) + theme_bw()` or replace the theme used in these graphs 
via `fg_replace_theme()`.   See accompanying vignette.


